---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

###This assignment takes data from a "personal activity monitoring device", such as a Fitbit or Jawbone, that was gathered on an anonymous individual during October and November 2012. It includes the number of steps taken in five minute intervals.
##Load Activity Data into R
```{r}
act_data<- read.csv("activity.csv", header = TRUE)
```
###View structure of data
```{r}
str(act_data)
```

###Load dependicies
```{r warning = FALSE, message = FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
```

###Convert data into Date format
```{r}
act_data$date <- ymd(act_data$date)
act_data <- tbl_df(act_data)
```


## What is mean total number of steps taken per day?
```{r}
by_day <- act_data %>% group_by(date) %>% summarise(total_by_day = sum(steps))
by_day
```

###Produce histogram of total steps per day
```{r}
hist(by_day$total_by_day / 1000, breaks = 15,
     col = "blue",
     xlab = "Steps Per Day (Thousand)", 
     ylab = "Number of Days",
     main = "Histogram of Total Steps
     Per Day")
```

###Find mean and median of total number of steps per day
```{r}
summary(by_day)
```


## What is the average daily activity pattern?
###Plot time series
```{r}
by_interval <- act_data %>% group_by(interval) %>% summarise(avg_steps = mean(steps, na.rm = TRUE))
## another way: by_int <- tapply(act_data$steps, act_data$interval, mean, na.rm = TRUE)
plot(avg_steps ~ interval, data = by_interval, type = "l",
     main = "Average Steps Per 5 Minute Interval Averaged Across All Days",
     xlab = "5 Minute Interval",
     ylab = "Average Steps")
     axis(side = 1, at = c(0, 500, 1000, 1500, 2000, 2355))
```

###Find interval with maximum average number of steps
```{r}
max_step_interval <- by_interval[which(by_interval$avg_steps == max(by_interval$avg_steps)),]
max_step_interval[[1]][1]
```


## Imputing missing values
###Calculate total in dataset
Since ```{r, eval = FALSE}summary(act_data)``` has previously shown NAs only appear in variable "steps",
```{r}
sum(is.na(act_data$steps))
```

###Find and replace NA with median NA of corresponding 5 minute interval
```{r}
replace_na <- function(act_data){
  median_steps_interval <-
    act_data %>% group_by(interval) %>% summarise(median_steps = median(steps, na.rm = TRUE))
  act_data2 <- act_data

  for(i in 1:nrow(act_data2)){
    if(is.na(act_data2$steps[i])){
      replace_na_with_index <- which(act_data2$interval[i] == median_steps_interval$interval)
      replace_na_with <- median_steps_interval[replace_na_with_index,]
      act_data2$steps[i] <- replace_na_with$median_steps
    }
  }
  act_data2
}

## create new dataset with NAs replaced
act_data_no_na <- replace_na(act_data)
```

###Create histogram of total steps per day with NAs replaced
```{r}
by_day_no_na <- act_data_no_na %>% group_by(date) %>% summarise(total_by_day = sum(steps))
hist(by_day_no_na$total_by_day / 1000, breaks = 15,
     col = "red",
     xlab = "Steps Per Day (Thousand)", 
     ylab = "Number of Days",
     main = "Histogram of Total Steps
     Per Day (NAs Replaced)")
```

###Compare to total steps per day with NAs removed.

* Mean and median total steps per day with NAs replaced
```{r}
mean_steps_na_replaced <- mean(by_day_no_na$total_by_day)
mean_steps_na_replaced
median_steps_na_replaced <- median(by_day_no_na$total_by_day)
median_steps_na_replaced
```

* Mean and median total steps per day with NAs removed
```{r}
mean_steps_no_na <- mean(by_day$total_by_day, na.rm = TRUE)
mean_steps_no_na
median_steps_no_na <- median(by_day$total_by_day, na.rm = TRUE)
median_steps_no_na
```

* Difference in mean and median
```{r}
mean_steps_na_replaced - mean_steps_no_na
median_steps_na_replaced - median_steps_no_na
```

Both mean and median are decreased when NAs are replaced with median steps by interval, instead of simply removing NA values.


## Are there differences in activity patterns between weekdays and weekends?
###Create variable indicating weekday or weekend (use NA replaced dataset)
```{r}
act_data_no_na <- act_data_no_na %>% mutate(weekday = wday(date))
in_week <- act_data_no_na$weekday %in% c(1:5)
act_data_no_na$weekday <- factor(in_week, labels = c("weekend", "weekday"))
```

###Create time-series plot facetted by weekend/weekday of total steps by interval
```{r}
int_grouped_act_data_no_na <- act_data_no_na %>% group_by(interval, weekday) %>% summarise(total_steps = sum(steps))

ggplot(int_grouped_act_data_no_na, aes(interval, total_steps)) +
  geom_line() +
  facet_wrap(~weekday, ncol = 1) + 
  labs(x = "Interval", y = "Number of steps")
```
Yes, there appears to be higher activity level during weekdays.
